Grove: A Bidirectionally Typed Collaborative Structure Editor Calculus

### 2024-10-25 OriginalSubmissionSource/introduction.tex
Updates from Overleaf

% Software development is a highly collaborative activity. 
Development teams typically collaborate with the aid of a version control system (VCS) like Git~\cite{git}, Subversion~\cite{subversion}, or Darcs~\cite{darcs}. 
These systems maintain a branching history of \emph{commits} to a source code repository, each consisting of a \emph{patch} together with various metadata, e.g. a human-readable commit message. Patches are imperative programs written in a 
\emph{patch language} defining a set of primitive editing commands.
The standard POSIX \li{patch} language, for example, specifies commands for inserting and deleting specified lines of text at specified line numbers within a file.

Developers do not typically express program edits using the patch language directly, nor in any case do version control systems typically have access to a log of the developer's edits. Instead, version control systems must \emph{synthesize} patches from the file system state using heuristic algorithms, such as the classic \li{diff} algorithm that synthesizes a patch that minimizes the edit distance between two file system states.\todo{cite}

When two patches, developed concurrently in branches based on a common ancestral commit, must be merged, version control systems deploy a \emph{three-way merge algorithm}. The standard approach is to apply the \emph{local patch} first, then modify the \emph{remote patch} by shifting its line numbers to account for the local patch's line insertions and deletions. This algorithm is an \emph{operational transform}~\cite{ot}. 
Character-level operational transforms are similarly used for real-time collaborative editing, e.g. in tools like Google Docs and in Visual Studio Code's Live Share feature~\cite{crdts-for-text}.

\paragraph{Problem 1: Merging Sensibly} 
When merging patches, conflicts (e.g. due to different modifications to the same location) are unavoidable. However, 
standard three-way merge algorithms for line-based patch languages commonly identify spurious conflicts, fail to identify legitimate conflicts, or silently duplicate or misplace code. Let us briefly review some classic problems with these systems. Appendix~\ref{appendix:merge-problems} shows examples of the problematic default behavior of Git in each of these scenarios.\todo{write this, can move to supplement if needed}

@@

These classic problems have motivated research into richer patch languages, more sophisticated patch synthesis algorithms, and corresponding improvements to three-way merge algorithms. 
For example, systems like Git address the nesting problem by allowing indentation changes from one patch to be merged with other changes that do not modify indentation. 
More sophisticated systems like GumTree, SmartDifferencer\todo{cites, others} deploy parsers and tree differencing algorithms to better
address the granularity problem.

@@

\paragraph{Contribution 1: Grove: A Collaborative Structure Editor Calculus} 
This paper considers the problem of collaborative editing for \emph{structure editors},
which eschew text editing. Instead, developers code by applying tree edit actions directly to a continuously evolving \emph{program sketch}, i.e. a syntax tree with holes, shown projected visually in various ways to the developer. 
Structure editing has been studied since the 1980s with the Cornell Program Synthesizer 
and research continues to this day, with numerous active projects including Scratch, Snap!, and other block-based editors (which are widely used in educational and end-user programming settings), Jetbrains MPS (which has been deployed in industry), and Hazel (a live functional programming environment rooted in a structure editor calculus called Hazelnut, which serves as 
an active research platform).

Inspired by Hazelnut, this paper introduces Grove, a \emph{collaborative structure editor calculus} for arbitrary syntax trees that does not suffer from the problems just outlined. 
This is in large part due to a substantial simplification of the overall collaborative
editing architecture. In particular, we \emph{eliminate patch synthesis (i.e. diff algorithms) entirely}, 
instead deriving patches directly from the log of edits performed by the developer. 
We also \emph{eliminate the need for three-way merge algorithms (i.e. operational transforms) entirely}.
Instead, we define the patch language such that all edits commute, 
so remote patches can be applied without transformation. 
We prove a \emph{convergence theorem} that ensures that replicas (a.k.a. clones) of a repository will converge to the same state 
when the same set of patches are applied, regardless of the order in which they are applied.
The patch language forms what is known as a \emph{commutative replicated data type (CmRDT)}.

@@

Structure editors are designed to provide editing affordances for trees, not graphs where there may be any number of children at a given location, vertices may have any number of parents, and where there may be cycles. 
To support conventional tree-based structure editing, 
we define a \emph{decomposition} of our program graph into a \emph{grove}, which is a set of 
programs with holes, local conflicts, and conflict references between them to account for motifs that arise when editing collaboratively. 
In particular, a location with no out-edges decomposes to a \emph{hole}. 
More than one out-edge at a given location decomposes to an explicitly represented \emph{local conflict}. 
More than one in-edge indicates that a vertex has a \emph{location conflict}, so we leave a \emph{location conflict reference} at each of the conflicting locations. 
Finally, cycles are broken during decomposition by leaving a \emph{cycle conflict reference} at an arbitrary (but deterministically chosen) edge. 
Resolving these various conflicts simply requires 
manipulating these constructs like any other syntactic construct, e.g. deleting all but one location conflict reference to determine a unique location for a node.

\paragraph{Problem 2: Semantic Gaps During Conflict Resolution}
When working with traditional version control systems,
resolving conflicts can take time and require reasoning about 
syntax, types, and program behavior.
Traditionally, however, conflicts are indicated by inserting extra-linguistic markers
into files. 
These markers are not typically understood by the parser, and 
because they include conflicting alternatives, they cannot generally be removed 
or concatenated to result in a sensible program. 
Consequently, language services that require a well-formed, meaningful program
(e.g. type error localization, go-to-definition, live evaluation and so on) 
either fail to operate or exhibit gaps in service, e.g. because they are relying on 
data from a compile prior to the merge attempt. Developers are left to reason without the aid of much of their tooling during conflict resolution. This is an instance of the more general \textbf{semantic gap problem} when programming tools encounter incomplete programs~\cite{HazelSNAPL}.

The previous work on the Hazelnut structure editor calculus addresses the semantic gap problem in the single-user setting by 
defining a type system and type error localization system (the \emph{marked lambda calculus}) for incomplete programs, i.e. programs with holes. 
Notably, type error localization is proven to be \emph{total}, i.e. the system is able to assign static meaning to \emph{every} syntactically well-formed expression by inserting \emph{marks} to localize errors. Marking employs local type inference as codified by a \emph{bidirectional type system} as well as \emph{gradual typing}, i.e. the theory of \emph{type holes}, to recover from situations where type errors make it impossible to 
determine a known type. 
A separate type hole filling phase deploys unification-based (i.e. non-local) type inference to fill type holes when possible, 
or allows the user to interactively select from hole fillings that partially satisfy generated constraints when there are type conflicts.

@@

This paper extends this work to develop a total type error localization system for groves, introduced in Contribution 1 above as the result of decomposing a commutatively edited graph into a set of terms with empty holes, local conflicts, location conflict references, and cycle conflict references. This paper develops a type (and type error localization) discipline for handling these novel constructs. We follow the marked lambda calculus in rooting our \emph{marked grove calculus} in bidirectional type checking, deploying gradual typing when conflicts do not allow a single type to be inferred, and then layering on a unification-based type inference system to opportunistically fill holes or suggest partial solutions when there are conflicting types due to conflicting syntax.

@@

\autoref{sec:Grove By Example} introduces Grove by example, demonstrating its behavior in each of the problematic scenarios named above. 
\autoref{sec:Formalism} then formally defines Grove's graph structure, commutative patch language, grove decomposition procedure, and edit action language. We establish key metatheoretic properties using the Agda proof assistant. 
\autoref{sec:Type System} instantiates Grove with a simply typed lambda calculus, then defines a bidirectional type and type localization system for groves and proves totality and other key metatheoretic properties using the Agda proof assistant. 
\autoref{sec:Implementation} describes our OCaml implementation of Grove, which is defined modularly to allow it to be instantiated with arbitrary syntax trees. 
\autoref{sec:Related Work} reviews related work in more detail. 
\autoref{sec:Discussion and Conclusion} concludes with a discussion and directions for future work on collaborative editing.

>>>>

% Software development is a highly collaborative activity. 
Development teams typically collaborate with the aid of a version control system (VCS) like Git, Subversion, or Darcs~\cite{DBLP:conf/haskell/Roundy05}. 
These systems maintain a branching history of \emph{commits} to a source code repository, each consisting of a \emph{patch} together with various metadata, e.g. a human-readable commit message. Patches are imperative programs written in a 
\emph{patch language} defining a set of primitive editing commands.
The standard POSIX \li{patch} language, for example, specifies commands for inserting and deleting specified lines of text at specified line numbers within a file.

Developers do not typically express program edits using the patch language directly, nor in any case do version control systems typically have access to a log of the developer's edits. Instead, version control systems must \emph{synthesize} patches from the file system state using heuristic algorithms, such as the classic \li{diff} algorithm that synthesizes a patch that minimizes the edit distance between two file system states \cite{DiffAlgorithm}.

When two patches, developed concurrently in branches based on a common ancestral commit, must be merged, version control systems deploy a \emph{three-way merge algorithm}. The standard approach is to apply the \emph{local patch} first, then modify the \emph{remote patch} by shifting its line numbers to account for the local patch's line insertions and deletions. This algorithm is an \emph{operational transform}~\cite{DBLP:conf/sigmod/EllisG89}. 
Character-level operational transforms are similarly used for real-time collaborative editing, e.g. in tools like Google Docs and in Visual Studio Code's Live Share feature.

\paragraph{Problem 1: Merging Sensibly} 
When merging patches, conflicts (e.g. due to different modifications to the same location) are unavoidable. However, 
standard three-way merge algorithms for line-based patch languages commonly identify spurious conflicts, fail to identify legitimate conflicts, or silently duplicate or misplace code. Let us briefly review some classic problems with these systems. 
%Appendix~\ref{appendix:merge-problems} shows examples of the problematic default behavior of Git in each of these scenarios.\todo{write this, can move to supplement if needed}

@@

These classic problems have motivated research into richer patch languages, more sophisticated patch synthesis algorithms, and corresponding improvements to three-way merge algorithms. 
For example, systems like Git address the nesting problem by allowing indentation changes from one patch to be merged with other changes that do not modify indentation. 
More sophisticated systems deploy parsers and tree differencing algorithms~\cite{DBLP:conf/sigmod/ChawatheG97, DBLP:journals/tse/FluriWPG07,DBLP:conf/kbse/FalleriMBMM14,DBLP:conf/doceng/Lindholm04,DBLP:conf/fase/NguyenNPN10,DBLP:journals/scp/SchwagerlUW15} to better
address the granularity problem.

@@

\paragraph{Contribution 1: Grove: A Collaborative Structure Editor Calculus} 
This paper considers the problem of collaborative editing for \emph{structure editors},
which eschew text editing. Instead, developers code by applying tree edit actions directly to a continuously evolving \emph{program sketch}, i.e. a syntax tree with holes, shown projected visually in various ways to the developer. 
Structure editing has been studied since the 1980s with the Cornell Program Synthesizer~\cite{DBLP:journals/cacm/TeitelbaumR81}
and research continues to this day, with numerous active projects including Scratch~\cite{maloney2010scratch}, Snap!, and other block-based editors (which are widely used in educational and end-user programming settings), Jetbrains MPS (which has been deployed in industry)~\cite{voelter2011language}, and Hazel (a live functional programming environment rooted in a structure editor calculus called Hazelnut, which serves as 
an active research platform)~\cite{DBLP:conf/popl/OmarVHAH17}.

Inspired by Hazelnut, this paper introduces Grove, a \emph{collaborative structure editor calculus} for arbitrary syntax trees that does not suffer from the problems just outlined. 
This is in large part due to a substantial simplification of the overall collaborative
editing architecture. In particular, we \emph{eliminate patch synthesis (i.e. diff algorithms) entirely}, 
instead deriving patches directly from the log of edits performed by the developer. 
We also \emph{eliminate the need for three-way merge algorithms (i.e. operational transforms) entirely}.
Instead, we define the patch language such that all edits commute, 
so remote patches can be applied without transformation. 
We prove a \emph{convergence theorem} that ensures that branches of a repository will converge to the same state 
when the same set of patches are applied, regardless of the order in which they are applied.
The patch language forms what is known as a \emph{commutative replicated data type (CmRDT)}~\cite{preguicca2018conflict,shapiro2011conflict}.

@@

Structure editors are designed to provide editing affordances for trees, not graphs where there may be any number of children at a given location, vertices may have any number of parents, and where there may be cycles. 
To support conventional tree-based structure editing, 
we define a \emph{decomposition} of our program graph into a \emph{grove}, which is a set of 
programs with holes, local conflicts, and conflict references between them to account for motifs that arise when editing collaboratively. 
In particular, a location with no out-edges decomposes to a \emph{hole}. 
More than one out-edge at a given location decomposes to an explicitly represented \emph{local conflict}. 
More than one in-edge indicates that a vertex has a \emph{relocation conflict}, so we leave a \emph{relocation conflict reference} at each of the conflicting locations. 
Finally, cycles are broken during decomposition by leaving a \emph{unicycle conflict reference} at an arbitrary (but deterministically chosen) edge. 
Resolving these various conflicts simply requires 
manipulating these constructs like any other syntactic construct, e.g. deleting all but one relocation conflict reference to determine a unique location for a node.

\paragraph{Problem 2: Semantic Gaps During Conflict Resolution}
When working with traditional version control systems,
resolving conflicts can take time and require reasoning about 
syntax, types, and program behavior.
Traditionally, however, conflicts are indicated by inserting extra-linguistic markers
into files. 
These markers are not typically understood by the parser, and 
because they include conflicting alternatives, they cannot generally be removed 
or concatenated to result in a sensible program. 
Consequently, language services that require a well-formed, meaningful program
(e.g. type error localization, go-to-definition, live evaluation and so on) 
either fail to operate or exhibit gaps in service, e.g. because they are relying on 
data from a compile prior to the merge attempt. Developers are left to reason without the aid of much of their tooling during conflict resolution. This is an instance of the more general \textbf{semantic gap problem} when programming tools encounter incomplete programs~\cite{DBLP:conf/snapl/OmarVHSGAH17}.

The previous work on the Hazelnut structure editor calculus addresses the semantic gap problem in the single-user setting by 
defining a type system and type error localization system (the \emph{marked lambda calculus}) for incomplete programs, i.e. programs with holes~\cite{DBLP:journals/pacmpl/ZhaoMDBPO24}. 
Notably, type error localization is proven to be \emph{total}, i.e. the system is able to assign static meaning to \emph{every} syntactically well-formed expression by inserting \emph{marks} to localize errors. Marking employs local type inference as codified by a \emph{bidirectional type system}~\cite{dunfield2019} as well as \emph{gradual typing}~\cite{siek2015}, i.e. the theory of \emph{type holes}, to recover from situations where type errors make it impossible to 
determine a known type. 
A separate type hole filling phase deploys unification-based (i.e. non-local) type inference to fill type holes when possible, 
or allows the user to interactively select from hole fillings that partially satisfy generated constraints when there are type conflicts.

@@

This paper extends this work to develop a total type error localization system for groves, introduced in Contribution 1 above as the result of decomposing a commutatively edited graph into a set of terms with empty holes, local conflicts, relocation conflict references, and unicycle conflict references. This paper develops a type (and type error localization) discipline for handling these novel constructs. We follow the marked lambda calculus in rooting our \emph{marked grove calculus} in bidirectional type checking, deploying gradual typing when conflicts do not allow a single type to be inferred, and then layering on a unification-based type inference system to opportunistically fill holes or suggest partial solutions when there are conflicting types due to conflicting syntax.

@@

\autoref{sec:Grove By Example} introduces Grove by example, demonstrating its behavior in each of the problematic scenarios named above. 
\autoref{sec:Formalism} then formally defines Grove's graph structure, commutative patch language, grove decomposition procedure, and edit action language. We establish key metatheoretic properties using the Agda proof assistant. 
\autoref{sec:Implementation} describes our implementation of the Grove Workbench, which is defined modularly to allow it to be instantiated with arbitrary syntax trees. 
\autoref{sec:Marking System} instantiates Grove with a simply typed lambda calculus, then defines a bidirectional type and type localization system for groves and proves totality and other key metatheoretic properties using the Agda proof assistant. 
\autoref{sec:Related Work} reviews related work in more detail. 
\autoref{sec:Discussion and Conclusion} concludes with a discussion and directions for future work on collaborative editing.

### 2024-10-25 OriginalSubmissionSource/related-work.tex
Updates from Overleaf

Contrast with a traditional diff

Similar to how git merged across moves.
A notable difference though, is that \texttt{git} detects
file moves by structural similarity rather than
our nominal equality.
(We exploit this in Subsection~REF:TODO).

Compare this to what happens when there are conflicts in a traditional,
line/diff-based version control system (e.g., \texttt{git}, \texttt{mercurial}, \texttt{svn}, etc.).
Those systems annotate files with difference markers (e.g., sequences of~\texttt{<},~\texttt{>}, or~\texttt{=})
when a merge conflict happens.
These show alternate versions of the code, and it is up to the user
to replace those alternates with the desired merged result.

Another place our model is different
is that changes are explicit rather than being inferred by a \texttt{diff} algorithm.
Finally, in our model, since all changes are in terms of
the structure of the code rather than lines,
all changes respect that structure.
This is unlike the line-based model where there is no guarantee
that a merge conflict follows the grammatical structure of the code.

\subsection{Trees}

\subsubsection{Structure editors}

Hazel~\citep{Omar:2019:10.1145/3290327}

\subsubsection{Tree diff}

@@

\subsection{Version Control}

@@

\subsection{Collaborative Editing}

@@

\subsection{Synchronization}

@@

\subsection{Homotopical Patch Theory}

@@

>>>>

Collaboration requires defining a patch language, a method for synthesizing patches from user actions, and an approach for merging patches.

\subsection{Patch Languages}
There have been many of different designs for patch languages and indeed many imperative data structures and their associated operations can be construed as patch languages in the most general sense. In the context of collaborative coding, patch languages can differ in the data structures they operate on (e.g. line-based text, character-based text, tree-structured data). 
They can also differ in how they identify locations within the data (e.g. by using numeric offsets, unique identifiers, or paths through a tree).
Finally, patch languages differ in which specific actions are supported explicitly. Insertion and deletion are common, while code relocation, copying, undo, and other operations are variously also included.

In this paper, our focus was on syntax trees with holes (i.e. program sketches~\cite{DBLP:conf/aplas/Solar-Lezama09, DBLP:conf/popl/OmarVHAH17}) and explicit conflicts, which we represented as directed graphs. We identify locations using unique IDs. We have a two-level patch language, with a low level graph patch language supporting only edge insertion and deletion and a higher-level user edit action language focused on insertion, deletion, and relocation, with some additional narrative consideration of copying and undo (a fuller account of which we leave to future work).  Our user edit language therefore forms a structure editor calculus, inspired closely by recent work on the Hazelnut structure editor calculus (which did not support relocation)~\cite{DBLP:conf/popl/OmarVHAH17} and patch languages for other tree-based data structures~\cite{DBLP:conf/sigmod/ChawatheG97, DBLP:journals/tse/FluriWPG07,DBLP:conf/kbse/FalleriMBMM14,DBLP:conf/doceng/Lindholm04,DBLP:conf/fase/NguyenNPN10,DBLP:journals/scp/SchwagerlUW15}.

\subsection{Patch Synthesis}
The most common approach to patch synthesis is to use a differencing algorithm to compare two states, e.g. from the file system, to generate a patch. 
The classic \li{diff} algorithm~\cite{DiffAlgorithm}, for example, minimizes edit distance for a patch language involving line insertions and deletions.

In the setting of tree-based editing, there have been a number of tree differencing algorithms described in the literature \cite{DBLP:conf/esa/Klein98,DBLP:journals/tcs/Bille05,DBLP:journals/talg/DemaineMRW09,DBLP:journals/fuin/AratsuHK10,DBLP:conf/sigmod/ChawatheG97, DBLP:journals/tse/FluriWPG07,DBLP:conf/kbse/FalleriMBMM14,DBLP:conf/doceng/Lindholm04,DBLP:conf/fase/NguyenNPN10,DBLP:journals/scp/SchwagerlUW15}. 
As described in \autoref{sec:Introduction}, synthesizing insertions and deletions is well-understood, but synthesizing relocations is more complex and requires heuristics.

The approach we explore in this paper is far simpler: we directly translate from the log of user edit actions to graph patches, without needing a differencing algorithm at all. This is only possible with a structure editor integrated with the version control system, but the benefit of direct visibility into the edit action log is that we do not need heuristics to synthesize relocations.

\subsection{Merging}
\subsubsection{Operational Transforms}
The most common approach to merging concurrently developed patches is to deploy an operational transform~\cite{DBLP:conf/sigmod/EllisG89} whereby locations in a remote patch are modified based on the action of a local patch. Standard three-way merge algorithms in version control systems deploy this approach, as do real-time collaborative editors. There have been a number of papers studying the algebraic properties of merging patches in this style.

For example, the Darcs~\cite{DBLP:conf/haskell/Roundy05} and Pijul version control systems, like Grove, represent repositories using sets of patches. However, unlike Grove, not all orderings of patches are algebraically sensible (e.g. commutative and associative), so the theory of these systems define and algebraically characterize when operational transforms do satisfy these properties. Recent work on homotopical patch theory \cite{DBLP:journals/jfp/AngiuliMLH16} has similarly developed an abstract algebraic framework for distinguishing sensible merges.

\subsubsection{CRDTs}

The observation that commutativity is a particularly helpful property when dealing with concurrent systems, including version control systems, has led to the development of a number of data structures centered on commutativity. These are known as CRDTs, which stands variously for \emph{conflict-free}, \emph{convergent} (CvRDT), or \emph{commutative replicated datatypes} (CmRDT) depending on particular details about the operations and the state representation~\cite{shapiro2011conflict}.
Our approach draws directly from this line of work: the Hazel patch language forms a CmRDT on directed graphs, from which we observe that we can derive a CmRDT for trees with explicit conflicts.

There have been other recent efforts to develop CmRDTs for tree data structures~\cite{DBLP:conf/icdcs/PreguicaMSL09, Highly-Available}. However, the focus in this work has been on avoiding conflicts and cycles entirely by applying \emph{ad hoc} heuristics for conflict resolution at merge-time, e.g. using reported timestamps or favoring particular directions in the tree. Our approach instead embraces manual conflict resolution, as is common practice in software projects where arbitrarily losing code is not acceptable.

% \url{https://www.waitingforcode.com/big-data-algorithms/conflict-free-replicated-data-types-flags-graphs-maps/read}

@@

% \subsection{Version Control}

@@

% \subsection{Collaborative Editing}

@@

% \subsection{Synchronization}

@@

% \subsection{Homotopical Patch Theory}

@@

% - diff
%     - A file comparison program 1985
%     - myers an O(ND) difference algorithm and its variations
% - merging
%     - a state-of-the-art survey on software merging
% - homotopy patch theory
%     - 7 10 15 16 28 31
%     - general algebraic framework for discussing patch languages
% - darcs
%     - unordered series of patches (but dependencies between them)
%     - we don't focus on inversion in this paper
%     - some properties of darcs patch theory
%         - http://urchin.earth.li/darcs/ganesh/darcs-patch-theory/theory/formal.pdf
%     - A formalization of darcs patch theory using inverse
% semigroups
%         - 2, 3, 5, 11, 18
%         - 15 = selective undo https://ww3.math.ucla.edu/camreport/cam09-83.pdf
%         - patch commutation requires an operational transform! https://ww3.math.ucla.edu/camreport/cam09-83.pdf
%         - (we also define it as a semigroup)
% - dagit - type correct changes
%     - Type-correct changes—a safe approach to version control
% implementation
%     - MS thesis
%     - Darcs: Distributed version management in haskell.
%     - "patches that modify different parts of the code are considered, by default, independent"
%     - patience diff
% - A categorical theory of patches
% - OT = "Concurrency control in groupware systems"
%     - Formal design and
% verification of operational transformation algorithms for copies convergence
% - The Semantics of Version Control
%     - https://dl.acm.org/doi/abs/10.1145/2661136.2661137?casa_token=hKqrcG9g7_UAAAAA:vN0l9AJhdnjc_2wnAIBl3TumxJBgy_4JYYSsG7522fYnbdfzh7Mpc0PIwUKcsX2VbwA0M67Cy_Y
%     - cite for this is a version control system
%     - CACM article by o'sullivan
%         - https://dl.acm.org/doi/pdf/10.1145/1562164.1562183
%         - "We are not by any means near the end
% of the road in the evolution of revisioncontrol systems. The field has received
% only fitful attention from academia.
% Much work could be done on its formal foundations, which could lead to
% more powerful and safer ways for developers to work together. Alas, I know
% of only one notable publication on the
% topic in the past decade.1
%  As a simple
% example, when merging potentially
% conflicting changes, almost everybody
% uses either three-way merging, which
% is decades old, or unpublished ad hoc
% approaches in which there is little reason to be confident."
% - touchdevelop
%     - https://dl.acm.org/doi/pdf/10.1145/2846661.2846672
% - tree CRDTs
%     - A Highly-Available Move Operation for Replicated Trees
%         - "last writer wins" conflict resolution not explicit conflict representation
%         - using timestamps
%         - Martin et al 28, 29 CRDT for XML data + Kleppman and Beresford CRDT for JSON but no moves
%         - [31] outline approaches to handling conflicts on trees but no algor
%         - treedoc 34 also no move
%         - 37, 38, 39 Ts for trees
%         - 40 defines OT with a move operation but need a total order broadcast (not commutative)
%     - Maram
%         - The price to pay is that some move operations “lose”, i.e., have no
% effect; achieving the same end result as previous correct approaches but at a lower cost
%         - https://arxiv.org/pdf/2103.04828
%     - UDR tree -- ordering
%     - Najafzadeh et al Subtree_1 requires locks
%     - Tao et al [17]
%         - . Merging semantics for conflict updates in
% geo-distributed file systems
%         - Can lead to duplication
%         - 
% - tree diff based approaches
%     - https://dl.acm.org/doi/10.1007/978-3-662-54494-5_9
%         - 14, 15, 21, 24, 25
%     - Change distilling: tree differencing
% for fine-grained source code change extraction.
%     - Fine-grained
% and accurate source code differencing.
%     - Tree-based Version Control in Envision
%     - A three-way merge for XML documents
%     - Operation-Based, Fine-Grained Version Control Model for Tree-Based Representation
%     - https://www.sciencedirect.com/science/article/pii/S0167642315000532
% - do not guess!
%     - https://dl.acm.org/doi/pdf/10.1145/2508075.2508092 (brief note)
%         - Cedalion: a language for language-oriented programming
% - variability aware execution
%     - https://dl.acm.org/doi/abs/10.1145/2786805.2803208
% - text CRDTs?

### 2024-10-25 figures/figure-cycle.tex
Updates from Overleaf

%\hfill
\hskip0.07\columnwidth

@@

\begin{subfigure}[t]{0.2\textwidth}

@@

\path (-9em,0cm) node (node/MultiParent/ex) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:MultiParent:e}}}};

@@

\path [draw,->,alice step] (node/MultiParent/ex) to[star] (-2em, 0cm |- node/Cycle/a);

@@

    "\vMultiCycleTimes" [> "\eMultiCycleTimes"', >alice edge, alice node],
    "\vMultiCyclePlus" [> "\eMultiCyclePlus", >alice edge, alice node]

@@

\begin{subfigure}[t]{0.2\textwidth}

@@

\figureCode{\parens{\hole{} * \hole{}} + $\conflictHole{\multiVertex{\vidMultiCycleTimes},\multiVertex{\vidMultiCyclePlus}}$%

@@

\caption{Example of Cycles.}

>>>>

% \hskip0.07\columnwidth

@@

\begin{subfigure}[t]{0.25\textwidth}

@@

% \path (-9em,0cm) node (node/MultiParent/ex) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:MultiParent:e}}}};

@@

% \path [draw,->,alice step] (node/MultiParent/ex) to[star] (-2em, 0cm |- node/Cycle/a);

@@

    "\vMultiCycleTimes" [> "\eMultiCycleTimes"'],
    "\vMultiCyclePlus" [> "\eMultiCyclePlus"]

@@

\begin{subfigure}[t]{0.18\textwidth}

@@

\figureCode{\parens{\hole{} * \hole{}} + $\binaryConflictHole{\multiVertex{\vidMultiCycleTimes}}{\multiVertex{\vidMultiCyclePlus}}$%

@@

\caption{(a) We start with a tree with a multiplication,  \vMultiCycleTimes{}, and addition, \vMultiCyclePlus{}, at the leaves. (b) Alice relocates them both, such that \vMultiCycleTimes{} is the parent of \vMultiCyclePlus{}. (c) Bob relocates them both, such that \vMultiCyclePlus{} is the parent of the \vMultiCycleTimes{}. (d) In the merged state, there is a cycle in the graph. Because the terms have a common parent, there is a local conflict. Because the cycle is connected to the rest of the graph, the cycle is broken during decomposition by relocation conflict references as shown.}
